---
title: 'Lab 6: Change (Pre-Post Test)'
author: "Wednesday Bushong"
date: "3/24/2017"
output: pdf_document
---

Today's lab will be short and sweet :)

## Overview of Pre-Post Test Analyses

In a repeated measures design, instead of having only one observation per subject, you have 2 or more. This is indeed the norm in cognitive psychology! Today we'll just be tackling the case where we have a pre-test and post-test value. We have 2 options for how we want to analyze change between a pre- and post-test:

* **Analyze straight change**: Here, we compute the change in the dependent variable between time A and time B.

* **Analyze residual change**: Here, we predict the value at time B from the value at time A. If this is positive, it means there was positive change over time; if negative, negative over time. If there's an effect of the independent variable, then 

We also have a choice for whether we want to use regression or ANOVA -- if we're using an ANOVA, we'll want to include an error term for participant.

## The Data 

Females and males had 'handicap' scores taken at two times. We're interested in whether gender affects the timecourse of these values.

```{r, warning = FALSE}
# Load libraries & data
library(foreign)
library(tidyr)
library(ggplot2)

# data
d <- read.spss("data.sav", to.data.frame = TRUE)
# what does the data look like?
summary(d)
head(d)
# label participants
d$participant <- as.factor(1:nrow(d))
```

There's already a change score that has been computed for us in the data. Let's make sure that these are the right values: 

```{r}
all(d$change == d$fhndicap - d$bhndicap) # all function tells you if this statement returns all TRUEs 
```

## First, Some Plotting

We can visualize the straight change scores or the scores over time.

```{r}
p.change <- ggplot(d, aes(x = bgender, y = change)) +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange")
p.change

d.gathered <- gather(d, time, value, bhndicap:fhndicap)
# we 'gather' the data so that bhndicap and fhndicap are compressed into one column 
# that columns essentially represents time, so I call it 'time' (argument 2 of gather)
# the other column, which I assign the name 'value', will contain the actual values that were the original contents of the bhndicap and fhndicap columns

# plot data over time
p.over.time <- ggplot(d.gathered, aes(x = time, y = value, 
	group = bgender, shape = bgender)) +
	stat_summary(fun.y = mean, geom = "point", size = 4) +
	stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.1) +
	stat_summary(fun.y = mean, geom = "line")
p.over.time
```

## Straight Change

Super easy: just predict change scores from gender! We can either predict straight change, or predict value from the time * gender interaction. 

```{r}
m <- aov(change ~ bgender + Error(participant), d) # Error(participant) adds in the within-subject error term into the model
summary(m)

# equivalent to:
m.int <- aov(value ~ bgender * time + Error(participant), d.gathered)
summary(m.int)

m2 <- lm(change ~ bgender, d)
summary(m2)
```

## Residual Change

For a residual change analysis, we predict scores at the second time point (fhndicap) from the first time point & our independent variable, gender. 

```{r}
m3 <- aov(fhndicap ~ bhndicap + bgender, d)
summary(m3)

m4 <- lm(fhndicap ~ bhndicap + bgender, d)
summary(m4)
```

